Keep your Laziness in Check

### 2018-02-25 paper/main.tex
[paper] more work on section 2

\leo{These examples should go in the next section. Putting them here to avoid conflicts with Hengchu}

\leo{Example 1: Okasaki queues}

For the second example, we turn to a different data functional data
structure where both a strict and a lazy variant can be situationally
useful: maps. There are two reasonable laziness behaviors that
implement a map interface; we could implement a completely strict map,
or we could lazily evaluate the values stored as needed.
\footnote{A third option could be to have lazy keys, but they would
necessarily be forced at every insertion or lookup, with no gain.}
Haskell's standard library provides these two variants in the form of
\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
to {\em specify} the difference between the two versions! At best, one
could hope for a hand-crafted suite of unit tests containing bottoms
to observe a difference in termination.
\leo{One thing we could do is ignore the knapsack and expand on the above.}

Moreover, it is important to be able to distinguish whether such a
datatype follows to a desired laziness behavior. For instance, the
standard functional solution for dynamic programming problems
crucially uses laziness. In particular, consider the famous knapsack
problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.
%
The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

solution :: Map (Int, Int) Int
solution = fromList [ ((j, k),
                       if (j-1) < 0 || k - ws !! j < 0
                       then 0
                       else max (solution ! (j-1, k))
                                (solution ! (j-1, k - ws !! j) + vs !! j)
                      )
                    | j <- [0..n-1], k <- [0..w] ]
\end{inlinecode}

\leo{Is this true? Where would the infinite loop be? there is a bottom to the recursion.
I thought the problem was that it computes everything when it doesn't need to}
Using the lazy variant of \lk{Map} here is critical for the correctness
of \lk{solution}.
However, using the lazy variant of \lk{Map} here is critical for the
correctness of \lk{solution}. If we used a map strict in its values
instead, then \lk{fromList} would have eagerly evaluated the \lk{if}
expression. The \lk{if} expression contains value-level recursion to
the solution map itself, and this would manifest as an infinite loop
at runtime. This is an incredibly subtle bug related to laziness.

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
%
\begin{figure}[H]
\centering
\begin{inlinecode}
data ListD a = NilD
             | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
\caption{\lk{ListD} type}
\label{fig:ListD}
\end{figure}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and a nullary constructor \lk{T}
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor
from \lk{Thunk}. For example, given the list \lk{[1, 2, 3]}, the
value \lk{ConsD T (ConsD T T)} represents the \textit{demand} that
reveals the first 2 \lk{Cons} constructors of the list, but does not
evaluate any other parts of the list. To improve readability of demand
values, we will also use the syntax \lk{[\_, \_, ...]} where
appropriate.

@@

Putting these pieces together, we arrive at an appropriate type for
the specification of \lk{rotate}: \lk{ListD a -> [a] -> [a] -> [a] ->
(ListD a, ListD a, ListD a)}.
%
\hzh{The description here needs improvement}
To specify the laziness behavior of \lk{rotate}, we need to inspect
the 4 different cases of \lk{rotate}. When both \lk{fs} and \lk{bs}
are empty, the accumulator is returned directly; when \lk{fs} is an
empty list, \lk{rotate} immediately recurses on the tail of \lk{bs};
when \lk{bs} is empty, \lk{rotate} produces a new list with the
original head of \lk{fs} as the new head, and the tail of that list is
the result of a recursion; finally, when both \lk{fs} and \lk{bs} are
not empty, \lk{rotate} also immediately produces a new list with the
appropriate head, and conses \hzh{is this a word} the original head
of \lk{bs} onto the accumulator.

Turning our intuition above into a Haskell function, we complete the
specification of \lk{rotate} in \Cref{fig:rotate-spec}
\begin{figure}[H]
\centering
\begin{inlinecode}
rotate_spec :: ListD a
            -> [a] -> [a] -> [a]
            -> (Thunk (ListD a), Thunk (ListD a), Thunk (ListD a))
rotate_spec NilD                     fs bs as = (E NilD, E NilD, E NilD)
rotate_spec dOut@(ConsD dElem dTail) fs bs as =
  case (fs, bs) of
    ([],           []) ->
      (E NilD, E NilD, E dOut)
    ([],      b : bs') ->
      let (_, dBs', dBAs) = rotate_spec dTail [] bs' (b:as)
      in case dBAs of
           T                -> (E NilD, T, T)
           E (ConsD dB dAs) -> (E NilD, ConsD dB dBs', dAs)
           E NilD           -> error ``Impossible''
    (_ : fs',      []) ->
      let (dFs', _, dAs) = rotate_spec dTail fs' [] as
      in  (E (ConsD dElem dFs'), E NilD, dAs)
    (_ : fs', b : bs') ->
      let (dFs', dBs', dBAs) = rotate_spec dTail as' bs' (b:as)
          dFs                = E (ConsD dElem dFs')
      in case dBAs of
           T                -> (dFs, T, T)
           E (ConsD dB dAs) -> (dFs, E (ConsD dB dBs'), dAs)
           E NilD           -> error ``Impossible''
\end{inlinecode}
\caption{Specification for \lk{rotate}}
\label{fig:rotate-spec}
\end{figure}

>>>>

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
%
\begin{figure}[H]
\centering
\begin{inlinecode}
data ListD a = NilD
             | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
\caption{\lk{ListD} type}
\label{fig:ListD}
\end{figure}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and a nullary constructor \lk{T}
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor
from \lk{Thunk}. For example, given the list \lk{[1, 2, 3]}, the
value \lk{ConsD T (ConsD T T)} represents the \textit{demand} that
reveals the first 2 \lk{Cons} constructors of the list, but does not
evaluate any other parts of the list. To improve readability of demand
values, we will also use the syntax \lk{\_ : \_ : \_} where
appropriate.

@@

Putting these pieces together, we arrive at an appropriate type for
the specification of \lk{rotate}: \lk{ListD a -> [a] -> [a] -> [a] ->
(ListD a, ListD a, ListD a)}.
%
\hzh{The description here needs improvement}
To specify the laziness behavior of \lk{rotate}, we return to Okasaki's insight:
as long as the front list is longer than the back list, demanding the first
element from the result of \lk{rotate} only causes evaluation of the first
element of the front list ~\cite{Okasaki95simpleand}. Using this precondition
for \lk{rotate}, we arrive at the following executable
specification \lk{rotate\_spec}.

Turning our intuition above into a Haskell function, we complete the
specification of \lk{rotate} in \Cref{fig:rotate-spec}
\begin{figure}[H]
\centering
\hzh{TODO}
\begin{inlinecode}
TODO: rewrite this spec so it looks like the invariant in Okasaki's paper.
\end{inlinecode}
\caption{Specification for \lk{rotate}}
\label{fig:rotate-spec}
\end{figure}

\lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example. \hzh{should we show what a
counter example looks like?}

For the second example, we turn to a different functional data structure where
both a strict and a lazy variant can be situationally useful: maps. There are
two reasonable laziness behaviors that implement a map interface; we could
implement a completely strict map, or we could lazily evaluate the values stored
as needed.
\footnote{A third option could be to have lazy keys, but they would
necessarily be forced at every insertion or lookup, with no gain.}
Haskell's standard library provides these two variants in the form of
\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
to {\em specify} the difference between the two versions! At best, one
could hope for a hand-crafted suite of unit tests containing bottoms
to observe a difference in termination.
\leo{One thing we could do is ignore the knapsack and expand on the above.}

Moreover, it is important to be able to distinguish whether such a datatype
follows a desired laziness behavior. For instance, the standard functional
solution for dynamic programming problems crucially uses laziness \hzh{is this
solution really standard? Is there some work we can cite to establish this
approach as the canonical approach for dynamic programming in functional
programming?}. In particular, consider the famous knapsack problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.
%
The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

solutionPre :: Map (Int, Int) Int -> Map (Int, Int) Int
solutionPre soln = fromList [ ((j, k),
                               if (j-1) < 0 || k - ws !! j < 0
                               then if ws !! j <= k then vs !! j else 0
                               else max (soln ! (j-1, k))
                                        (soln ! (j-1, k - ws !! j) + vs !! j)
                              )
                            | j <- [0..2], k <- [0..50] ]

solution :: Map (Int, Int) Int
solution = fix solutionPre
\end{inlinecode}

However, using the lazy variant of \lk{Map} here is critical for the
correctness of \lk{solution}. If we used a map strict in its values
instead, then \lk{fromList} would have eagerly evaluated the \lk{if}
expression. The \lk{if} expression contains value-level recursion to
the solution map itself, and this would manifest as an infinite loop
at runtime. This is an incredibly subtle bug related to laziness.

To catch this behavior through testing, we can write another specification that
captures this correctness condition. The well-foundedness \hzh{is this a word?}
of \lk{solution} depends on what is evaluated when the value at key \lk{(j, k)}
of the map is demanded. Let $\lk{solutionPre}_n$ be the $n$-times iterated
pre-fixpoint \lk{solutionPre . solutionPre . ... . solutionPre}, then if we
demand the value at \lk{(j, k)} from the result of $\lk{solutionPre}_n$, its
input map should not be evaluated at the \lk{(j, k)} entry since otherwise
evaluating the value at \lk{(j, k)} becomes an infinite loop at the fixpoint.

We can write another executable specification for this property
of $\lk{solutionPre}_n$ (\Cref{fig:spec-map}). Testing the implementation based on
strict maps also immediately produces a counter example.

\begin{figure}[H]
\centering
\hzh{TODO: specification for map}
\begin{inlinecode}
\end{inlinecode}
\caption{Specification for \lk{solutionPre}}
\label{fig:spec-map}
\end{figure}
